Detecting confounder and center bias in machine learning models

### 2021-10-28 manuscript.tex

The tests provide a strict control for Type-I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously partly unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.

@@

In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (demonstrated on Fig. \ref{fig:sim-h0-demo}). These conditions are very common for case of predictive modelling \citep{garcia2009study, kristensen2017whole} (Supplementary Material \ref{sup:nomlinviol})

@@

In the following section, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
\subsubsection*{The partial confounder test}

@@

In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g. \citep{bohning1992multinomial}).

@@

Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in our case:

@@

The simulated predicted values are constructed analogously:

$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$

@@

To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing, its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq. \ref{eq:linreg}, as described by \cite{berrett2020conditional}) with the following simulation parameters:
\begin{itemize}
    \item sample size: $n = 1000$
    \item $w_{c\hat{y}} = 0$ (H0 simulations only)
    \item all combinations of:
    \begin{itemize}
        \item $w_{yc} \in \{0.5, 1, 2, 3\}$ and
        \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
    \end{itemize}
    \item non-normality ($f_{\delta = 0.1, \epsilon = 2)}$ and non-linearity (sigmoid $g$)
\end{itemize}

For each parameter combination, 1000 repetitions were performed and false positive rate was calculated as the ratio of p-values smaller than $\alpha = 0.05$.

These simulation cases are visualized (with $w_{yc} = w_{y\hat{y}} = 2$) on the left of Figure \ref{fig:sim-h0-demo}.

@@

Both for the partial and the full confound test, 100 repetitions were performed of all combination of the following parameter values: 
\begin{itemize}
    \item $w_{yc} \in \{0.5, 1, 2, 3\}$
    \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
    \item $w_{c\hat{y}} \in \{0, 0.2, 0.4, 0.6\}$
    \item $n \in \{50, 100, 500, 1000\}$
    \item linear and sigmoid dependence
    \item normal conditional distribution
\end{itemize}

The 'partial' and 'full' confound tests, as implemented in version 0.20.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.

All analysis code used for the simulations is available on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}}.

@@

\subsubsection*{HCP: testing age- and acquisition batch-bias in fluid intelligence prediction}

Human Connectome Project contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}. Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices \citep{duncan2000neural}.

@@

The total of 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge-regression \citep{hoerl1970ridge}. The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

@@

The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs. HC) with a L2-regularized logistic regression (as e.g. in \citep{dadi2019benchmarking}). The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type-I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
  Type-I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row). Non-normal conditional distribution on the second plot is illustrated with blue density diagrams and dependency is depicted with blue dotted lines an all plots. False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values. The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065. Values below this threshold indicate a valid type-I error control and colored white.
  }
  \label{fig:sim-h0-demo}
\end{figure}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\yhat}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\yhat}$ (rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%. First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for the type I error rates is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

The conditional permutation based \emph{partial} and \emph{full confounder tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.

@@

\begin{figure}[!b]
  \centering

@@

Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-full}, for the partial test on categorical data and the full tests on both numeric and categorical data). 
In case of a valid test, the expected type I error rate in the null hypothesis simulation cases (first row and first column of the figures) is less then or equal to the alpha level (0.05). The upper 95\% confidence limit for the type I error rates for one simulation case (i.e. a single tile on the figures) is 0.11 (binomial confidence intervals with 100 repetitions). 
The results suggest, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).

Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$). The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, from normal (Fisher's kurtosis: 0, skewness: 0) to severely skewed and heavy tailed (kurtosis: 42, skewness: -6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, in terms of type I error rate, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

Power estimates for the 'partial' test (with normality) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only 0.4-0.6\% to the variance of the predictions can already be successfully detected ($w_c = 0.05 - 0.07$) with a power of 60-90\%. With a sample size of 500, a confounder contributing to only 1.4-3\% to the variance of the predictions can be detected ($w_c = 0.1 - 0.14$) with a power greater than 80\% in most of the simulation cases. A sample size of 100 yields an approximate power of 60-100\% for a confounder-variance of 6-7\% ($w_c = 0.2 - 0.21$) in most of the cases. Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 15\% of variance, it is almost certainly detected with $n \geq 50$.

Violation of the normality assumptions for $\y$ and $\c$ has a negative effect on the power of the tests, however the loss of power is low or even negligible, if normality is not extremely violated (first 4 distributions on Fig. \ref{fig:sim-non-normal}A) and the confounder-to-signal ratio is greater or equal then 0.3 (Fig. \ref{fig:sim-non-normal}C).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

@@

After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig. \ref{fig:hcp}), neither for acquisition batch nor for age. Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD). The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001). While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig. \ref{fig:abide}). Several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.169$ and $0.111$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.

Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).

When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder successfully detected the resulting strong bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).

@@

The full confounder test was highly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$) Indicating that this model might have been (almost) exclusively driven by motion artifacts.

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different tests, probing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem allows placing assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized and computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}. It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained. Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.

>>>>

The tests provide a strict control for Type I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously partly unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.

@@

In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g. \citep{korn1984ranges} or Fig. \ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.

@@

In the following section, CPT is adapted for partial confounder testing (option 3). For an overview, see Fig. \ref{fig:overview}. The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
\subsubsection*{The partial confounder test}

@@

In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g. \citep{bennett1966multiple, jones1975proability}).

@@

Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced in this case) or e.g. a sigmoid-shaped function, in our case:

@@

The simulated predicted values are constructed similarly:

$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{y\hat{y}} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$

@@

To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing (Fig. \ref{fig:sim-h0-demo}), its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq. \ref{eq:linreg}, as described by \cite{berrett2020conditional}) with the following simulation parameters:
\begin{itemize}
    \item sample size: $n = 1000$
    \item $w_{c\hat{y}} = 0$ (i.e. H0 simulations only)
    \item all combinations of:
    \begin{itemize}
        \item $w_{yc} \in \{0.5, 1, 2, 3\}$ and
        \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
    \end{itemize}
    \item non-normality ($f_{\delta = 0.1, \epsilon = 2}$) and non-linearity (sigmoid $g$)
\end{itemize}

For each parameter combination, 1000 repetitions were performed and false positive rates were calculated as the ratio of p-values smaller than $\alpha = 0.05$.

The simulation cases are exemplified (with $w_{yc} = w_{y\hat{y}} = 2$) on the left of Figure \ref{fig:sim-h0-demo}.

@@

Both for the partial and the full confounder test, 100 repetitions were performed of all combination of the following parameter values: 
\begin{itemize}
    \item $w_{yc} \in \{0.5, 1, 2, 3\}$
    \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
    \item $w_{c\hat{y}} \in \{0, 0.2, 0.4, 0.6\}$
    \item $n \in \{50, 100, 500, 1000\}$
    \item linear and sigmoid dependence
    \item normal conditional distribution
\end{itemize}

The partial and full confound tests, as implemented in version 0.20.1 of the package \emph{'mlconfound'} were run with default parameters (1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations) and by implying categorical variables, where needed.

All code used for the simulations is available on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}}.

@@

\subsubsection*{HCP: testing age and acquisition batch bias in fluid intelligence prediction}

The Human Connectome Project dataset contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}. Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices \citep{duncan2000neural}.

@@

Each of the 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) was independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge regression \citep{hoerl1970ridge}. The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

@@

The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs. HC) with a L2-regularized logistic regression (see rationale in \citep{dadi2019benchmarking}). The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% fig: overview
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
  \caption{\textbf{Overview of the proposed partial confoudner test}. \\
    The partial confoudner test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM). The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m). The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confoudner variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
  }
  \label{fig:overview}
\end{figure}

\subsection{The partial and full confoudner tests}

The proposed \emph{partial} and \emph{full confoudner tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}. The inner workings of the tests is summarized on Fig. \ref{fig:overview}. In short, the tests model the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) and then use a parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder or target variable, for the partial and full confounder test, respectively. As a result, the permuted variables comply with the GAM-based conditional distribution. The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. The strength of this approach lies in avoiding to place any assumptions on the conditional distributions of the model output. This property ensures valid model diagnostics even in cases where Pearson and Spearman partial correlation and many other methods fail (as shown on Fig. \ref{fig:sim-h0-demo}).

@@

As suggested by theory and shown by the simulations, both of the proposed tests provide a valid Type I error control (Fig. \ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs. \ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig. \ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6). With this the proposed tests overcome the limitations of partial Spearman correlation and linear conditional permutations, demonstrated on Fig. \ref{fig:sim-h0-demo}.

For detailed results, see also Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}.

%fig:sim-h0-demo
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row). Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1) and dependency of the mean is depicted with blue dotted lines an all plots. False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values. The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065. Values below this threshold indicate a valid type I error control and colored white.
  }
  \label{fig:sim-h0-demo}
\end{figure}

%fig:sim-normal
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type I error control and power of the partial confoudner test based on simulations with normal conditional distribution and linear dependencies.} \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%. First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

%fig:fig:sim-non-normal
\begin{figure}[!b]
  \centering

@@

@@

Power estimates for the 'partial' test (with normal and linear simulations) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%. With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%. Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.

Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition- and age-bias of fluid intelligence prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is given on the graphs via $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the partial confound test. P-values of the 'full' confound test were all less then 0.0001, i.e. the confounders did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

@@

After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounder bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig. \ref{fig:hcp}), neither for acquisition batch nor for age. Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD). The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001). While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the partial confound test. P-values of the full confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig. \ref{fig:abide}). Several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately associated ($R^2=0.169$ and $0.111$, respectively) to a lower probability for the participant to be assigned to the ASD group by the model.

Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).

When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder test successfully detected the resulting strong bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).

@@

The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model might have been (almost) exclusively driven by motion artifacts.

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. The problem of testing confounder-bias in predictive models can considered as assessing \emph{conditional independence} between the predicted values, the the target and the confounder variable.
Importantly, in predictive modelling, the model predictions are often characterized by a non-normal and/or non-linear conditional dependence on the target variable, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power. In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios. This is clearly demonstrated on Fig. \ref{fig:sim-h0-demo} in case of partial Spearman correlation. Although warnings about these issues were given from early on \citep{korn1984ranges} and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated within the predictive modelling community.

From the three involved variables, the most problematic one is clearly the prediction output, known to often display non-normal and non-linear conditional dependencies on the other two. Therefore, a method being distribution-free for this variable, but inputting custom distributional information about the other two variables might provide a sufficient flexibility required for predictive model diagnostics.
